---
title: "31_DataVisualization"
author: "Melody Lu"
date: "2025-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
library(tidyverse)
library(janitor)
library(scales)
library(sjPlot)
```

```{r lp}

source("1_datacleaning.R")
```
Below is some code for visualizations that contextualize question 1: How does gender or race correlate with primary success rates (winning primaries), and do those correlations differ by party or by office (Senate, House, Governor)?
```{r}
# Fit models
##formula <- as.formula(primary_win ~ party + gender + race + ideology + incumbent)
##model_all <- run_logistic_model(all_candidates, formula)

##Q1 VISUALIZATION


#  win rate by gender
ggplot(candidates_q1, aes(gender, primary_outcome)) +
  stat_summary(fun = mean, geom = "bar", fill = "grey") +
  labs(title = "Primary Win Rate by Gender",
       y = "Win Rate", x = "Gender") +
  theme_minimal()

# win rate by gender separated into parties
ggplot(candidates_q1, aes(gender, primary_outcome, fill = party)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  scale_fill_manual(values = c("Democrat" = "blue", "Republican" = "red")) +
  labs(title = "Win Rate by Gender and Party",
       y = "Win Rate", x = "Gender",
       fill = "Party") +
  theme_minimal()

# histogram of vote % by gender
ggplot(candidates_q1, aes(primary_percent, fill = gender)) +
  geom_histogram(
    bins = 20,
    alpha = 0.6,
    position = "identity",
    color = "black"      # outline so white doesn't disappear
  ) +
  scale_fill_manual(values = c(
    "Female" = "black",
    "Male" = "white",
    "Nonbinary" = "red"
  )) +
  labs(
    title = "Distribution of Primary Vote % by Gender",
    x = "Primary Vote %",
    y = "Count",
    fill = "Gender"
  ) +
  theme_minimal()

# win rate by race
ggplot(candidates_q1, aes(race, primary_outcome)) +
  stat_summary(fun = mean, geom = "bar", fill="orange") +
  scale_y_continuous(labels = scales::percent) +
  labs(title="Primary Win Rate by Race", y="Win Rate", x="Race") +
  theme_minimal()

# win rate by gender × race
ggplot(candidates_q1, aes(x = gender, y = primary_outcome, fill = race)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Primary Win Rate by Gender and Race",
       x = "Gender", y = "Win Rate", fill = "Race") +
  theme_minimal()

# win rate by gender × race seperated by party
ggplot(candidates_q1, aes(x = gender, y = primary_outcome, fill = race)) +
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  scale_y_continuous(labels = scales::percent) +
  facet_wrap(~ party) +
  labs(title = "Win Rate by Gender and Race, Separated by Party",
       x = "Gender", y = "Win Rate", fill = "Race") +
  theme_minimal()

# win rate by office
ggplot(candidates_q1, aes(office, primary_outcome)) +
  stat_summary(fun = mean, geom = "bar", fill = "steelblue") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Win Rate by Office", y = "Win Rate", x = "Office") +
  theme_minimal()


# scatter + logistic curve
ggplot(candidates_q1, aes(primary_percent, primary_outcome)) +
  geom_jitter(height=0.05, alpha=0.5) +
  stat_smooth(method="glm", method.args=list(family="binomial"), se=TRUE, color="red") +
  labs(title="Logistic Regression of Win Probability vs Primary %", 
       x="Primary Vote %", y="Win Probability") +
  theme_minimal()

# heatmap data
heatmap_q1 <- candidates_q1 %>%
  group_by(party, office, gender) %>%
  summarise(
    n_total = n(),
    success_rate = mean(primary_outcome, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_total >= 5)  

# plot heatmap
ggplot(heatmap_q1, aes(x = office, y = gender, fill = success_rate)) +
  geom_tile(color = "white", size = 1.2) +
  geom_text(
    aes(label = sprintf("%.0f%%\n(n=%d)", success_rate * 100, n_total)),
    color = "white", fontface = "bold", size = 3.5
  ) +
  facet_wrap(~ party) +
  scale_fill_gradient2(
    low = "#d73027",      # low = red
    mid = "#fee08b",      # mid = yellow
    high = "#1a9850",     # high = green
    midpoint = 0.35,
    name = "Win Rate"
  ) +
  labs(
    title = "Primary Win Rates by Gender × Office × Party",
    x = "Office",
    y = "Gender"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold", size = 12),
    plot.title = element_text(face = "bold", size = 14)
  )

# logistic regression: main effects
model_q1 <- glm(
  primary_outcome ~ gender + race + party + office + primary_percent,
  data = candidates_q1,
  family = "binomial"
)

# odds ratios main model
or_main <- exp(cbind(OR = coef(model_q1),
                     confint(model_q1)))
or_main
```
Ok now that we have created visualizations for question number 1, we will follow suite with question number 2: How does the proportion of overlap (ideological overlap between parties) in Congress change around periods of large shifts in primary dynamics (e.g. big wave years)?

```{r np}
## Q2 VISUALIZATIONS
#congressinal overlap over time
ggplot(pol, aes(year, overlap, color = chamber)) +
  geom_line(size = 1.2) +
  labs(title = "Ideological Overlap in Congress Over Time",
       x = "Year", y = "Overlap",
       color = "Chamber") +
  theme_minimal()

# ideological distance over time
ggplot(pol, aes(x = year, y = party_mean_diff_d1, color = chamber)) +
  geom_line(size = 1.2) +
  labs(
    title = "Party Ideological Distance Over Time",
    x = "Year",
    y = "Ideological Distance",
    color = "Chamber"
  ) +
  theme_minimal()
```
